---
title: "Probability"
subtitle: "Random Variables, Probability Distributions"
format: 
  revealjs:
    smaller: true
    footer: Stat 801A
execute:
  echo: true
---

## Sampling

A [Simple Random Sample]{style="color: magenta"} (SRS) of size $n$ from a finite population is a subset selected in such a way that every subset of n elements is equally likely to be the one selected.

## Random Variables

A [random variable]{style="color: magenta"} is a function defined on a numeric population. We often say that this function value is the value assumed by the random variable. A probability distribution specifies the probabilities associated with all possible values the random variable can assume.

## Example

If $X$ represents the number of RED $M\&Ms$ in a vending machine size
package, then perhaps [the probability distribution of
$X$]{style="color: magenta"} is something like the following:\

     $x$            5                6                7                8                9                10               11
  ---------- ---------------- ---------------- ---------------- ---------------- ---------------- ---------------- ----------------
   $P(X=x)$   $\frac{2}{28}$   $\frac{2}{28}$   $\frac{5}{28}$   $\frac{7}{28}$   $\frac{5}{28}$   $\frac{4}{28}$   $\frac{3}{28}$
   
---

```{r}
#| echo: false
#| eval: true

library(ggplot2)
x <- 5:11
tops <- c(2,2,5,7,5,4,3)/28
ggplot(mapping = aes(x = x, y = tops)) + 
  geom_col() + 
  labs(x = 'x', y = 'P(X=x)') + 
  theme_bw()
```

## Discrete Random Variables

When the elements of a population consists of a countable discrete set of values the random variable defined on that population is called discrete random variable.

Properties:

- $0\leq P(X=x)$ for all x
- $\sum_x P(X=x)=1$

## Continuous Random Variables

When the population consists of a set of values in a continuous range the random variable defined on that population is called continuous random variable. The probability distribution of a continuous random variable is a smooth curve, and the area under the curve between specified values of the random value correspond to a probability.

```{r}
#| fig-align: center
#| echo: false
x <- seq(-4,4, l = 1000)
ggplot(mapping = aes(x = x,
                     y = dnorm(x))) + 
  geom_path() + 
  theme_void()
```

## Continuous Random Variables

Properties:

- $0\leq f(x)$ for all $x$
- $\int f(x) dx=1$

## Continuous RV (example)

Suppose that our population consists of all numbers in the interval (0,2) occurring with various frequencies as specified by $f(x)$.

\
\
\
\
\


- $P(1/2\leq X\leq 1/2)=$

\
\
\

- $P(0\leq X\leq 2)=$

\
\
\


- $P(1\leq X\leq 2)=$

## Common Distributions

For some interesting populations that occur naturally, theoretical probability distributions and associated random variables, have been defined.

Discrete

- Binomial
- Poisson
- Negative Binomial

Continuous

- Normal
- Gamma
- Chi squared
- F 
- t


## Binomial Experiment

A binomial experiment consists of taking $n$ successive SRS, each of size 1, replacing elements between draws, and counting the number of "successes" (S) obtained, $m$. The alternative is "failures" (F). 

- $\pi=m/n$
- $1-\pi=(N-m)/N$

Let $X_i$ be the random variable that assumes the value of the outcome of the $i$-th draw for each $i=1,2,\dots,n$.

## Binomial Distribution

Let $X_i$ assume the value 1 if the outcome of the $i$-th draw is S or the value of 0 if the outcome is F. For a repeated process, $n$ random variables, $X_1,X_2,\dots,X_n$, are distributed as follows:

   $x$         0        1
---------- --------- -------
 $P(X=x)$   $1-\pi$   $\pi$


## Binomial Distribution (contd.)

Now define a new random variable $Y$ so that 

$$Y = X_1 + X_2 + X_3 + \cdots + X_n$$

- $Y$ represents the number of "successes" for the $n$ random variables
- $Y$ is said to be a binomial random variable.


## Binomial Distribution (contd.)

$$
Y\sim \text{Binomial}(n,\pi);\quad y=1,2,\dots,n; \quad 0\lt \pi\lt 1
$$

\
\
\

$P(Y=y)=$


## Mean and Variance of a Discrete Random Variable

- The [mean]{style="color: magenta"} of a discrete random variable $X$
  is: $$E(X) = \mu = \sum_x\,x \cdot P(X=x)$$

- The [variance]{style="color: magenta"} of a discrete random variable
  $X$ is: $$V(X) = \sigma^2= \sum_x\,(x - E(X))^2\cdot \ P(X=x)$$

- The [standard deviation]{style="color: magenta"} of a random variable
  $X$ is $$\sqrt{V(X)}=\sigma$$

## Mean and Variance of a Discrete Random Variable (contd.)

- Using the previous formula, the [mean]{style="color: magenta"} of a
  Binomial random variable $Y$, is shown to be $E(Y)=\mu = n \pi$

- The [variance]{style="color: magenta"} is shown to be
  $V(Y) =\sigma^2= n\pi(1 - \pi)$.

- Thus the [standard deviation]{style="color: magenta"} of a Binomial
  random variable is $\sigma= \sqrt{n\,\pi(1 - \pi)}$

## Binomial Distribution: Example

Consider a Binomial population where $Y\sim Bin(3,1/3)$. 

- Mean:

\
\
\

- Variance:

\
\
\

- Standard Deviation:

## Assumptions of the Binomial Model

1.  $n$ identical trials are made.

2.  Each trial results in either [Success($S$)]{style="color: green"} or
    [Failure($F$)]{style="color: red"}. (only two possible outcomes).

3.  The probability of success ($\pi$) remains the same from trial to
    trial.

4.  The trials are independent.

5.  Interest is in the number of successes obtained in the $n$ trials.
    The number is viewed as being the value assumed by a Binomial
    ($n,\ \pi$) random variable $Y$.
    
## Binomial Distribution: R

Let $Y\sim Bin(n,\pi)$

- $P(Y=y)$ is given by `dbinom(x=y, size=n, prob=pi)`
- $P(Y\leq y)$ is given by `pbinom(x=y, size=n, prob=pi)`
    
    
## Mean and Variance of a Continuous Random Variable

- The [mean]{style="color: magenta"} of a continuous random variable $X$
  is: $$E(X) = \mu = \int_x\,x \cdot f(x) dx$$

- The [variance]{style="color: magenta"} of a continuous random variable
  $X$ is: $$V(X) = \sigma^2= \int_x\,(x - E(X))^2\cdot \ f(x) dx=E(X^2)-E(X)^2$$

- The [standard deviation]{style="color: magenta"} of a random variable
  $X$ is $$\sqrt{V(X)}=\sigma$$
  
## Normal Random Variable

The Normal probability distribution is an example of a continuous distribution, and is defined in the interval $(-\infty,\infty)$. This is a bell-shaped distribution and is commonly used to model continuous data. 

\

$f(x)=$

\

where 

- $E(X)=$

\

- $V(X)=$

\

- $\sqrt{V(X)}$=

## Properties of the Normal Distribution

- Symmetric about the mean $\mu$
- Probabilities calculated as 

$$
P(a \leq X \leq b) = \frac{1}{\sqrt{2\pi}\sigma}\int_a^b e^{-\frac{1}{2}(\frac{x-\mu}{\sigma})^2} dx
$$

## Standard Normal Distribution

The standard normal distribution is a normal distribution with $\mu=0$ and $\sigma^2=1$, and is represented by the random variable $Z$. If $X\sim N(\mu,\sigma^2)$, then the following transformation converts $X$ into a standard normal random variable.

$$
Z=\frac{X-\mu}{\sigma}
$$

## Standard Normal Distribution

If $X\sim N(0,1)$, then we can use a Z-table to compute areas under the curve. For example,

$P(a\leq X\leq b)=$


## Standard Normal Distribution: R

We may also use R to compute probabilities for random variables that follow a normal distribution. While any $\mu$ and $\sigma$ can be selected, the default is $0$ and $1$, respectively.

- `dnorm()`: computes $f(z)$
- `pnorm()`: computes $p$ from $P(Z\leq z)=p$ for a specified $z$
- `qnorm()`: computes $z$ from $P(Z\leq z)=p$ for a specified $p$

## Standard Normal Distribution: Examples


1.  $P(Z \geq 0) = P(Z \leq 0) = 0.50$

```{r}
1-pnorm(0); pnorm(0)
```


2.  $P(0 \leq Z \leq 0.12) =P(Z<0.12)-P(Z<0)=0.5478-0.5 = 0.0478$

```{r}
pnorm(0.12) - pnorm(0)
```


4. $P(-0.12 \leq Z \leq 0.12) = 2 \cdot P(0 \leq Z \leq 0.12) = 0.0956$

```{r}
pnorm(0.12) - pnorm(-0.12)
```

## Standard Normal Distribution: Examples

1.  Find $z$ such that $P(0 \leq Z \leq z) = 0.4846$ 

```{r}
qnorm(p = (0.4846 + pnorm(0)))
```


2.  Find $z$ such that $P(z \leq Z \leq 0) = 0.2257$

```{r}
qnorm(p = -(0.2257 - pnorm(0)))
```


3.  Find $z$ such that $P(Z \geq z) = 0.011$

```{r}
qnorm(0.011, lower.tail = F)
```

## Normal Distribution

Similar to the previous problems, if we know $\mu$ and $\sigma$, we can solve probability problems using the `mean` and `sd` arguments in the `xnorm()` functions.

```{r}
qnorm(0.975, mean = 5, sd = 0.2)
```

## Sampling Distribution of the Sample Mean $\bar{Y}$

Define the Sample Mean as

$$
\bar{Y}=\frac1n\sum_{i=1}^n Y_i
$$

where $Y_1,Y_2,\dots,Y_n$ are taken independently from the population. That is, taking the sum of all of the values and dividing it by the count of values. Any function of $Y_1,Y_2,\dots,Y_n$ is a random variable, including $\bar{Y}$.

## Properties of the Sample Mean $\bar{Y}$

If $Y_1,Y_2,\dots,Y_n$ are each independently distributed from a normal distribution with mean $\mu$ and variance $\sigma^2$, then

- $E(\bar{Y})=\mu$
- $V(\bar{Y})=\sigma^2/n$

Standard Error of the Mean: the standard deviation of $\bar{Y}$


## Central Limit Theorem

When sample size $n$ is large, the random variable $\bar{Y}$ is approximately normally distributed with mean $\mu$ and variance $\sigma^2/n$. That is, 

$$
\bar{Y}\approx N(\mu,\sigma^2/n)
$$

## Central Limit Theorem: Simulation

Let's take $n=20$ and let $Y\sim N(50, 16)$. A single sample of size 20 from this population may look like the following:

```{r}
set.seed(313)
y <- rnorm(n = 20, mean = 50, sd = sqrt(16))
head(y)
mean(y)
var(y)
sd(y)
```

## Central Limit Theorem: Simulation

Let's say that we take many samples of size 20 from the population. We would expect $\bar{Y}$ to be approximately normal with mean 50 and variance 16/20 = 0.8.

- `replicate()`: repeat an expression n times (different n from `rnorm`)
- `apply()`: for each column (MARGIN = 2), apply the mean function

```{r}
many_samples <- replicate(n = 1000, rnorm(n = 20, mean = 50, sd = 3))
y_bar <- apply(many_samples, MARGIN = 2, mean)
hist(y_bar)
```

## Central Limit Theorem: Simulation

Let's say that we take many samples of size 20 from the population. We would expect $\bar{Y}$ to be approximately normal with mean 50 and variance 16/20 = 0.8.

- `replicate()`: repeat an expression n times (different n from `rnorm`)
- `apply()`: for each column (MARGIN = 2), apply the mean function

```{r}
mean(y_bar)
var(y_bar)
sd(y_bar)
```


## Central Limit Theorem: Example

Suppose that you are estimating the average weight of sheep in a large herd. You obtain a random sample of size $n=50$, and can assume that $\mu=40$ and $\sigma^2=4$. What is the probability that the sample mean will be within 0.1 pounds of $\mu$?

$$
\bar{Y}\approx 
$$

## Central Limit Theorem: Example

$$
P(39.9\leq \bar{Y} \leq 40.1)
$$
**Standard Normal Distribution**

```{r}
pnorm((40.1-40) / sqrt(4/50)) - pnorm((39.9-40) / sqrt(4/50))
```

**Normal Distribution**

```{r}
pnorm(40.1, mean = 40, sd = sqrt(4/50)) - pnorm(39.9, mean = 40, sd = sqrt(4/50)) 
```

## Poisson Distribution

Poisson distribution is used to model the number of occurrences of “rare” events in certain intervals of time and space. A Poisson random variable is a discrete random variable as it represents a count.

$$
P(X=x)=\frac{e^{-\lambda}\lambda^x}{x!}; \quad x=0,1,2,3,\dots
$$

where $\lambda$ is the mean parameter and represents the expected number of events occurring in the interval of interest.

## Poisson Distribution Examples

- $X$ = \# of alpha particles emitted from a polonium bar in an 8 minute
  period\

- $Y$ = \# of flaws on a standard size piece of manufactured product
  (e.g., 100m coaxial cable, 100 sq.meter plastic sheeting)\

- $Z$ = \# of hits on a web page in a 24h period

## Poisson Distribution Properties 

Let $X\sim Poi(\lambda)$.

- E(X) = 

\
\
\

- V(X) = 

\
\
\

- If $X_i\sim Poi(\lambda)$ for $i=1,2,\dots,n$, then $\sum_{i=1}^n X_i \sim Poi(n\lambda)$


## Poisson Distribution: R

Let $X\sim Poi(\lambda)$.

- $P(X = x)$: `dpois(x=x, lambda=lambda)`

\
\
\

- $P(X \leq x)$: `ppois(x=x, lambda=lambda)`


## Example problems

Let $\mu=10$ and $n=30$. When required, assume that $\sigma^2=4$. Compute each of the following problems for the Binomial, Normal, and Poisson distributions, as well as for $\bar{Y}$.

- $P(X\leq 4)$
- $P(X\lt 4)$
- $P(12\leq X \leq 24)$
- $x$ such that $P(X\leq x)=0.025$
- $x$ such that $P(X>x)=0.975$ 


## Extra Mean/Variance Properties

- $E(X)=\int_x x\cdotf(x) dx$

\

- $E(aX + B) = aE(X)+b$

\

- $V(X)=E((X-\mu)^2)=E(X^2)-E(X)^2$

\

- $V(aX + B)=a^2V(X)$ 



