---
title: "Project 1 Solutions"
format: html
---

# Problem 1

Read in the data and print the first 10 observations.

```{r}
ne_work_commute <- read.csv("travel-time-31109.csv")
head(ne_work_commute, n=10)
```


# Problem 2

Use `dim()` to determine the dimensions of the dataset.

```{r}
dim(ne_work_commute)
```

The dataset contains `r nrow(ne_work_commute)` rows and `r ncol(ne_work_commute)` columns.

# Problem 3

Compute the mean of `TRANTIME`.

```{r}
mean(ne_work_commute$TRANTIME)
```

The mean travel time is `r round(mean(ne_work_commute$TRANTIME), 2)` minutes.

# Problem 4

Compute the median of `TRANTIME`.

```{r}
median(ne_work_commute$TRANTIME)
```

The median travel time is `r round(median(ne_work_commute$TRANTIME), 2)` minutes.

# Problem 5

Compute the variance of `TRANTIME`.

```{r}
var(ne_work_commute$TRANTIME)
```

The variance of travel time is `r round(var(ne_work_commute$TRANTIME), 2)` minutes squared.

Variance measures the amount of variability in travel time using squared deviations from the mean.

# Problem 6

Compute the standard deviation of `TRANTIME`.

```{r}
sd(ne_work_commute$TRANTIME)
```

The standard deviation of travel time is `r round(sd(ne_work_commute$TRANTIME), 2)` minutes.

# Problem 7

Use `summary()` on `TRANTIME`.

```{r}
summary(ne_work_commute$TRANTIME)
```

The values returned by `summary()` have the following interpretations:

- **Minimum:** The smallest observed travel time.
- **1st Quartile:** The value below which approximately 25% of observations fall.
- **Median:** The middle travel time.
- **Mean:** The arithmetic average travel time.
- **3rd Quartile:** The value below which approximately 75% of observations fall.
- **Maximum:** The largest observed travel time.

# Problem 8

Create a histogram of `TRANTIME`.

```{r}
hist(
  ne_work_commute$TRANTIME,
  main = "Distribution of Travel Time to Work",
  xlab = "Travel Time (minutes)",
  ylab = "Number of Respondents"
)
```

The data is right-skewed with large outliers.

# Problem 9

Would the mean or median be a better description of travel time?

The median would be a better description since there are large values that would increase the mean.

# Problem 10

Create `filtered_travel_time` by removing outliers using the IQR method.

An observation is considered an outlier if

$$
x < Q_1 - 1.5(IQR)
$$

or

$$
x > Q_3 + 1.5(IQR).
$$

First calculate the quartiles and IQR.

```{r}
ne_summary <- summary(ne_work_commute$TRANTIME)

iqr <- IQR(ne_work_commute$TRANTIME)

q1 <- as.numeric(ne_summary[2])
q3 <- as.numeric(ne_summary[5])
iqr
```

Calculate the lower and upper boundaries:

```{r}
lower_bound <- q1 - 1.5 * iqr
upper_bound <- q3 + 1.5 * iqr

lower_bound
upper_bound
```

Now retain only observations within these boundaries:

```{r}
filtered_travel_time <- ne_work_commute$TRANTIME[
  ne_work_commute$TRANTIME >= lower_bound &
  ne_work_commute$TRANTIME <= upper_bound
]
```


# Problem 11

Create a histogram of `filtered_travel_time`.

```{r}
hist(
  filtered_travel_time,
  main = "Travel Time to Work After Removing Outliers",
  xlab = "Travel Time (minutes)",
  ylab = "Number of Respondents"
)
```

After removing the outliers, the data is somewhat symmetrical with some right-skew.

# Problem 12

Repeat Problems 3–6 using `filtered_travel_time` and arrange the results into a formatted table.

```{r}
filtered_summary <- data.frame(
  Statistic = c(
    "Mean",
    "Median",
    "Variance",
    "Standard Deviation"
  ),
  Value = c(
    mean(filtered_travel_time),
    median(filtered_travel_time),
    var(filtered_travel_time),
    sd(filtered_travel_time)
  )
)

knitr::kable(
  filtered_summary,
  digits = 2,
  col.names = c("Statistic", "Value"),
  caption = "Summary Statistics After Removing Outliers"
)
```


# Problem 13

Describe what changed after removing the outliers.

Calculate the original and filtered summary statistics:

```{r}
original_stats <- c(
  Mean = mean(ne_work_commute$TRANTIME),
  Median = median(ne_work_commute$TRANTIME),
  Variance = var(ne_work_commute$TRANTIME),
  SD = sd(ne_work_commute$TRANTIME)
)

filtered_stats <- c(
  Mean = mean(filtered_travel_time),
  Median = median(filtered_travel_time),
  Variance = var(filtered_travel_time),
  SD = sd(filtered_travel_time)
)

rbind(
  Original = original_stats,
  Filtered = filtered_stats
)
```

The median stayed the same. The mean slightly decreased, and the variance / standard deviation greatly decreased.

# Problem 14

Let $X$ be normally distributed with the mean and variance calculated from `filtered_travel_time`.

Thus,

$$
X \sim N(\mu,\sigma^2),
$$

where $\mu$ and $\sigma^2$ are the mean and variance of `filtered_travel_time`.

```{r}
mu <- mean(filtered_travel_time)
sigma <- sd(filtered_travel_time)

mu
sigma
```

Note that `pnorm()` requires the **standard deviation**, not the variance, so `sigma` is supplied to the `sd` argument.

## (a) Probability of less than 10 minutes

```{r}
p_less_10 <- pnorm(
  10,
  mean = mu,
  sd = sigma
)

p_less_10
```


## (b) Probability of between 10 and 20 minutes

We want

$$
P(10 < X < 20).
$$

Using cumulative probabilities,

$$
P(10 < X < 20)
=
P(X < 20)-P(X < 10).
$$

```{r}
p_10_20 <- pnorm(
  20,
  mean = mu,
  sd = sigma
) -
  pnorm(
    10,
    mean = mu,
    sd = sigma
  )

p_10_20
```

## (c) Probability of greater than 20 minutes

We want

$$
P(X > 20).
$$

Using the complement,

$$
P(X > 20)=1-P(X\leq20).
$$

```{r}
p_greater_20 <- 1 - pnorm(
  20,
  mean = mu,
  sd = sigma
)

p_greater_20
```

