---
title: "Project 1"
date: '2026-01-01'
author: STAT 801A, Fall 2026
date-format: "[Due on] MMMM D, YYYY at 9:00 PM"
format: 
  pdf:
    include-in-header:
      text: '\pagenumbering{gobble}'
---

## Overview

The American Community Survey is a large national survey that collects information on many aspects of life in the United States. You are tasked with analyzing a subset of the 2022 survey for people who live and work in Lancaster County, Nebraska. In this particular subset, respondents were asked how long it takes for them to commute to work and only responses for people who drive alone are included. The data is available in the file `travel-time-31109.csv`.

**Variable Descriptions:**

- `RESPONDENT`: Unique identifier for each respondent
- `TRANTIME`: Travel time to work in minutes
- `DEPARTS`: Time of day the respondent leaves for work (e.g., 7:30 AM = 730)
- `ARRIVES`: Time of day the respondent arrives at work (e.g., 7:30 AM = 730)

```{r}
#| eval: false
#| echo: false
library(tidyverse)
dat <- read.csv('usa_00008.csv')
ne <- dat %>% 
  filter(STATEFIP==31 & COUNTYFIP == 109) %>% # State and county
  filter(TRANWORK == 10) %>%  # Travel by car, truck, or van
  filter(CARPOOL == 1) %>%  # Does not carpool
  filter(RIDERS == 1) %>% # Verify one rider
  filter(PWSTATE2 == 31 & PWCOUNTY == 109) %>% # Work in Lancaster County, NE
  mutate(RESPONDENT = row_number()) %>% # Add a unique identifier for each respondent
  select(RESPONDENT, TRANTIME, DEPARTS, ARRIVES) # Select only the variables of interest
write_csv(ne, 'travel-time-31109.csv')

ne %>% 
  # slice_sample(n = 100) %>% 
  ggplot(mapping = aes(x = TRANTIME)) +
  geom_histogram()

mean(ne$TRANTIME)
median(ne$TRANTIME)
sd(ne$TRANTIME)
var(ne$TRANTIME)
summary(ne$TRANTIME)
hist(ne$TRANTIME)
```

## Problems

Complete each of the following problems; solutions and brief explanations are provided below. The solution code reads `travel-time-31109.csv` from the project folder.

```{r}
#| echo: false
# Load packages used in the solutions
library(tidyverse)
library(knitr)
```

## Solution: Problem 1

```{r}
#| echo: true
# Read the data and show the first 10 rows
ne_work_commute <- read.csv("travel-time-31109.csv")
head(ne_work_commute, 10)
```

## Solution: Problem 2

```{r}
#| echo: true
# Get dimensions and display a one-sentence summary
dim_ne <- dim(ne_work_commute)
dim_ne
paste0("The dataset has ", dim_ne[1], " rows and ", dim_ne[2], " columns.")
```

## Solution: Problem 3

```{r}
#| echo: true
# Mean of TRANTIME
mean_tran <- mean(ne_work_commute$TRANTIME, na.rm = TRUE)
mean_tran
```

## Solution: Problem 4

```{r}
#| echo: true
# Median of TRANTIME
median_tran <- median(ne_work_commute$TRANTIME, na.rm = TRUE)
median_tran
```

## Solution: Problem 5

```{r}
#| echo: true
# Variance of TRANTIME
var_tran <- var(ne_work_commute$TRANTIME, na.rm = TRUE)
var_tran
```

## Solution: Problem 6

```{r}
#| echo: true
# Standard deviation of TRANTIME
sd_tran <- sd(ne_work_commute$TRANTIME, na.rm = TRUE)
sd_tran
```

## Solution: Problem 7

```{r}
#| echo: true
# Quartiles / summary of TRANTIME
summary(ne_work_commute$TRANTIME)
```

## Solution: Problem 8

```{r}
#| echo: true
# Histogram of TRANTIME using ggplot2
ggplot(ne_work_commute, aes(x = TRANTIME)) +
  geom_histogram(binwidth = 5, color = "black", fill = "skyblue") +
  labs(title = "Histogram of Travel Time (minutes)", x = "TRANTIME (minutes)", y = "Count")
```

## Solution: Problem 9

```{r}
#| echo: true
# Compare mean and median and comment on which is preferable
mean_tran
median_tran
```

The mean is compared to the median above; if the mean is substantially larger than the median the distribution is right-skewed and the median is the more robust (preferred) measure of center.

## Solution: Problem 10

```{r}
#| echo: true
# Compute IQR-based bounds and create filtered_travel_time (non-outliers)
q1 <- quantile(ne_work_commute$TRANTIME, 0.25, na.rm = TRUE)
q3 <- quantile(ne_work_commute$TRANTIME, 0.75, na.rm = TRUE)
iqr_val <- IQR(ne_work_commute$TRANTIME, na.rm = TRUE)
lower <- q1 - 1.5 * iqr_val
upper <- q3 + 1.5 * iqr_val
filtered_travel_time <- ne_work_commute$TRANTIME[ne_work_commute$TRANTIME >= lower & ne_work_commute$TRANTIME <= upper]
length(filtered_travel_time)
head(filtered_travel_time)
```

## Solution: Problem 11

```{r}
#| echo: true
# Histogram of filtered_travel_time
filtered_df <- tibble(TRANTIME = filtered_travel_time)
ggplot(filtered_df, aes(x = TRANTIME)) +
  geom_histogram(binwidth = 5, color = "black", fill = "lightgreen") +
  labs(title = "Histogram of Filtered Travel Time (no IQR outliers)", x = "TRANTIME (minutes)", y = "Count")
```

## Solution: Problem 12

```{r}
#| echo: true
# Summary statistics for filtered_travel_time and formatted table
mean_f <- mean(filtered_travel_time, na.rm = TRUE)
median_f <- median(filtered_travel_time, na.rm = TRUE)
var_f <- var(filtered_travel_time, na.rm = TRUE)
sd_f <- sd(filtered_travel_time, na.rm = TRUE)

summary_table <- tibble(
  Statistic = c("Mean", "Median", "Variance", "SD"),
  Value = c(mean_f, median_f, var_f, sd_f)
) %>%
  mutate(Value = round(Value, 2))

knitr::kable(summary_table, caption = "Summary statistics for filtered_travel_time (rounded to 2 decimals)")
```

