---
title: "Project 2 Solutions"
format: html
---

# Problem 1

```{r}
z.test <- function(y, mu = 0, sigma = 1,
                   alternative = c('two.sided', 'less', 'greater'),
                   conf.level = 0.95) {
  
  alternative <- match.arg(alternative)
  message(paste('Number of missing values:', sum(is.na(y))))
  y_orig <- y
  y <- y[!is.na(y)]
  # Calculations here
  xbar <- mean(y)
  n <- length(y)
  se <- sigma / sqrt(n)
  z <- (xbar-mu)/se
  alpha <- 1-conf.level
  
  # Hypothesis tests
  if(alternative == "two.sided") {
    lcl <- xbar - qnorm(1-alpha/2)*se
    ucl <- xbar + qnorm(1-alpha/2)*se
    p.value <- 2*pnorm(abs(z), lower.tail = F)
  } else if (alternative == 'less') {
    lcl <- -Inf
    ucl <- xbar + qnorm(1-alpha)*se
    p.value <- pnorm(z)
  } else if (alternative == 'greater') {
    lcl <- xbar - qnorm(1-alpha)*se
    ucl <- Inf
    p.value <- pnorm(z, lower.tail = FALSE)
  } else {
    stop("Invalid alternative provided")
  }
  
  
  # Organize results
  results <- list(
    data = y_orig,                                      # Original data
    sample_mean = xbar,                                # Mean of the data
    standard_error = se,                             # Standard error
    z = z,                                          # Test statistic
    parameters = list(mu = mu, sigma = sigma),     # Parameters        
    alternative = alternative,                     # The alternative hypothesis
    alpha = alpha,                                      # Significance level
    confidence_interval = list(lcl = lcl, ucl = ucl),    # Confidence interval
    p.value = p.value                                     # p-value of the test
    
  )
  
  # Output list of results
  return(results)
}
```

# Problem 2

## Part a

```{r}
z.test.data <- read.csv('z-test-data.csv')
z.test(y=z.test.data$prob2a,
       mu=30, sigma=sd(z.test.data$prob2a),
       alternative = 'greater')
```

The p-value is less than 0.05, meaning that we reject the null hypothesis. There is evidence to conclude that the true mean is greater than 30. 

## Part b

```{r}
z.test(y=z.test.data$prob2b,
       mu=15, sigma=sd(z.test.data$prob2b),
       alternative = 'less')
```

The p-value is greater than 0.05, meaning that we fail to reject the null hypothesis. There is not evidence to conclude that the true mean is less than 15. 

## Part c

```{r}
z.test(y=z.test.data$prob2c,
       mu=50, sigma=sd(z.test.data$prob2c),
       alternative = 'two.sided')
```

The p-value is less than 0.05, meaning that we reject the null hypothesis. There is evidence to conclude that the true mean is different from 50. 

# Problem 3

```{r}
mn_river <- read.csv('mn-river-n-samples.csv')
y <- mn_river$N_mg_l
qqnorm(y); qqline(y)
hist(y)
boxplot(y)
```

For the part, values tend to follow the line in the Q-Q plot. The histogram shows a little bit of skewness, but nothing out of the ordinary. The boxplot looks fairly symmetrical. The data appears to be approximately normally distributed.

# Problem 4

$$
H_0: \mu=4.19
$$

$$
H_0: \mu>4.19
$$

# Problem 5

```{r}
# Remove missing values
y <- y[!is.na(y)]

# Calculate necessary values
n <- length(y)
xbar <- mean(y)
alpha <- 0.05
se <- sd(y)/sqrt(n)

test_statistic <- (xbar-4.19)/se
p_value <- pt(test_statistic, df = n-1, lower.tail = F)
ci <- xbar + c(-1,1)*qt(p=1-alpha, df=n-1)*se
knitr::kable(data.frame(
  test_statistic = test_statistic,
  p_value = p_value,
  lower_cl = ci[1],
  upper_cl = Inf
))
```


# Problem 6

```{r}
t.test(x=y, alternative = 'greater', mu = 4.19)
```


# Problem 7

The results of (5) and (6) are the same, but (6) is in a more structured format. 


# Problem 8

There is evidence to conclude that the true average Nitrogen (in mg/L) in the Minnesota River is greater than the previous historical average of 0.19 (p-value=0.02614). With 95% confidence, the true average quantity of Nitrogen is greater than 4.201 mg/L.

