---
title: Inferences about $\mu$
subtitle: "Part 2: Unknown $\\sigma^2$, t-distribution"
footer: STAT 801A
format: 
  revealjs:
    smaller: true
    incremental: false
execute: 
  echo: true
---

::: frame
## Inferences About when $\sigma$ is unknown

- For large sample sizes, it follows from the CLT that $\bar{Y}$ is approximately Normally distributed i.e., $\bar{Y} \dot{\sim}
  N(\mu,\frac{\sigma^2}{n})$.

- It follows that the random variable $$T_{n-1} =\Big( \bar{Y} - \mu \Big) / \Big( S/\sqrt{n} \Big)$$ has approximately the [Student's $t$]{style="color: magenta"} [distribution]{style="color: magenta"} with $n-1$ [degrees of freedom]{style="color: magenta"}.

- Here $Y_1, Y_2, \ldots,Y_n$ are sampling random variables and $\bar{Y} =
  \frac{1}{n} \displaystyle\sum_i Y_i$ is thus a random variable.
:::

::: frame
## Inferences About when $\sigma$ is unknown - cont.

The denominator of $T_{n-1}$ is the random variable 

$$
S = \sqrt{\frac{1}{n-1} \sum_{i=1}^n (Y_i - \bar{Y})^2}
$$

- If sampling from a Normal distribution, the sample mean $\bar{Y}$ has a Normal distribution (exactly) and therefore $(\bar{Y}-\mu)/(S/\sqrt{n})$ will have a Student's $t$ distribution (exactly), regardless of sample size.
:::

::: frame
## 

Until now, we have assumed that $n$ is large and $\sigma$ is known, or $n$ is large and $n$ is large enough also to use $s$ as an approximation to $\sigma$.

- Since we used the CLT to derive confidence limits and tests of hypotheses, the exact nature of the sampled population was not required to be specified.

- Now we require that the population distribution to be Normal. We don't need to know $\sigma$ or have a large sample, i.e. $n$ to be large.

- In this situation, for any $n$, ($\bar{Y} - \mu)/(S/\sqrt{n})$ will have a Student's t-distribution with d.f.$= n-1.$

- Using this fact we can obtain confidence intervals and conduct tests of hypotheses even though $\sigma$ is unknown.


:::

## Standard Normal vs. Student's t-distribution

::: frame
## Properties of the Student's distribution

There are many t-distributions each specified by a single parameter called [degrees of freedom ($df$)]{style="color: magenta"}.

- Like the standard normal population, the distribution is [symmetric]{style="color: magenta"} about 0 and has mean equal to 0.

- The t-distribution has variance $df/(df-1)$, and hence is more variable than the standard normal distribution which has variance equal to 1.

- We say that the t-distribution has [heavier tails]{style="color: magenta"} than the standard normal distribution.

- As the degrees of freedom $df$ increases, the t-distribution approaches that of the standard normal distribution.

:::


:::: frame
## 

The $t$ table gives quantiles 0.90, 0.95, 0.975, 0.99, 0.995, and 0.999 for $t$ distributions with selected *df*.

If $\alpha$ = 0.05, then $t_\alpha$ satisfies $P(T_{10} > t_{.05}) = 0.05$ and from the t table , $t_{.05} = 1.812$. If $\alpha$ = 0.05, then $t_{\alpha/2} \, \equiv t_{0.025}= 2.228$\
[Comparison of Normal and $t$ Quantiles]{style="color: magenta"}

::: center
+:---------+-----------:+--------:+--------:+--------:+--------:+--------:+--------:+
| $t_\alpha$ with indicated df                                                      |
+----------+------------+---------+---------+---------+---------+---------+---------+
| $\alpha$ | $z_\alpha$ | 10      | 20      | 30      | 40      | 60      | 240     |
+----------+------------+---------+---------+---------+---------+---------+---------+
| 0.1      | 1.28       | 1.37    | 1.32    | 1.31    | 1.30    | 1.30    | 1.28    |
+----------+------------+---------+---------+---------+---------+---------+---------+
| 0.05     | 1.65       | 1.81    | 1.72    | 1.70    | 1.68    | 1.67    | 1.65    |
+----------+------------+---------+---------+---------+---------+---------+---------+
| 0.01     | 2.33       | 2.76    | 2.53    | 2.46    | 2.42    | 2.39    | 2.34    |
+----------+------------+---------+---------+---------+---------+---------+---------+
:::
::::

## t-distribution R Functions

The process of obtaining $t_\alpha$ is nearly identical to that of $z_\alpha$. Again, recall that R computes $t_\alpha$ such that $P(T>t_\alpha)=\alpha$. For example, we can obtain $t_{0.05}$ with the following code:

```{r}
alpha = 0.05; n = 20
qt(1-alpha, df = n-1)
```



Or we can compute $P(T<1)$ with 10 $df$ using the following:

```{r}
pt(q=1, df=10)
```



::: frame
## Confidence Interval for $\mu$ based on the t-distribution

Let the sample size be $n$, and for $\alpha$ be a specified value, say, for e.g. $.05$.

- A $(1-\alpha)100\%$ confidence interval for $\mu$ is given by $$\left(\bar{y} - t_{\alpha/2} \, \frac{s}{\sqrt{n}}, \quad
          \bar{y} + t_{\alpha/2} \, \frac{s}{\sqrt{n}}\right)$$

- This may also be written as $\left(\bar{y} - t_{\alpha/2} s_{\bar{y}}, \quad \bar{y} +
           t_{\alpha/2} s_{\bar{y}} \right)$

- Here, $t_{\alpha/2}$ is the $1-\alpha/2$ percentile of the t-distribution with $n-1$ degrees of freedom and $s_{\bar{y}}$ is the [standard error of the mean]{style="color: magenta"}.
:::

:::: frame
## 


::::

::: frame
## Test Procedures based on the t-distribution:

[Hypotheses:]{style="color: magenta"}\
[Case 1:]{style="color: red"}$H_0: \mu \leq \mu_0\ \mbox{vs.}\ H_a: \mu >\mu_0$ (right-tailed test)\
[Case 2:]{style="color: red"} $H_0: \mu \geq \mu_0\ \mbox{vs.}\ H_a: \mu <\mu_0$ (left-tailed test)\
[Case 3]{style="color: red"} $H_0: \mu = \mu_0\ \mbox{vs.}\ H_a: \mu \not =\mu_0$ (two-tailed test)\
[T.S:]{style="color: magenta"} $t_c  =  \frac{\bar{y}-\mu_0}{s/\sqrt{n}}$\
[R.R:]{style="color: magenta"} For Type I error probability of $\alpha$:\
[Case 1:]{style="color: red"} Reject $H_0:$ if $t \geq t_{\alpha,(n-1)}$\
[Case 2:]{style="color: red"} Reject $H_0:$ if $t \leq -t_{\alpha,(n-1)}$\
[Case 3:]{style="color: red"} Reject $H_0:$ if $|t| \geq t_{\alpha/2,(n-1)}$
:::

::: frame
## Test Procedures based on the t-distribution - cont.:

[Level of Significance (p-value):]{style="color: magenta"}\
[Case 1:]{style="color: red"} $p = P(T_{n-1} > t_c$).\
[Case 2:]{style="color: red"} $p = P(T_{n-1} < t_c$).\
[Case 3:]{style="color: red"} $p= 2P(T_{n-1} > |t_c|)$.\

:::

## Exercise

A massive multistate outbreak of food-borne illness was attributed to *Salmonella enteritidis*. Epidimiologists determined that the source of the illness was ice cream. They sampled nine production runs from the company that produced the ice cream to determine the level of *Salmonella enteritidis* in the ice cream.

These levels (MPN/g) are as follows\
$.593\quad .142\quad .329\quad .691\quad .231\quad .793\quad .519\quad .392\quad .418$\
Use the data to determine whether the mean level of *Salmonella enteritidis* in the ice cream is greater than .3 MPN/g with $\alpha=.01$\

::: frame
## Exercise - cont.

[Solution:]{style="color: magenta"}\
Need to test $H_0: \mu \leq .3$ vs. $H_a: \mu > .3$

From the data $\bar{y}=.456$ and $s= .2128$ are computed, giving $$t_c=  \frac{\bar{y}-\mu_0}{s/\sqrt{n}}=\frac{.456-.3}{.2128/\sqrt{9}}=2.21$$ Because for the one-tailed test we need $t_{\alpha,(n-1)}$; we look up $t_{.01}$ with $df=9-1=8$. It is $2.896$. Thus the rejection region is: $t>2.896$

:::



::: frame
## Exercise - cont.

Since $t_c= 2.214$ does not exceed $2.896$ it is not in the R.R.\
Thus, there is insufficient evidence in the data to reject $H_0$ i.e to say that the mean level of *Salmonella enteritidis* exceeds the dangerous level of .3 MPN/g.\
The [p-value]{style="color: magenta"} to be computed is $P(T_8>2.21)$\
To calculate this exactly using the t-table is not possible since it is tabulated for only a few values of $a$.\
However, we can bound the p-value by noting that, for $(df=8)$, 2.21 lies between 1.86 and 2.306.\
This gives [$.025<\mbox{p-value}<.05$]{style="color: magenta"} showing that the [p-value is not less than our $\alpha$ of $.01$]{style="color: magenta"}. Thus we fail to reject $H_0$.
:::

## Exercise - R code

Recall that we are testing $H_0:\mu=0.3$ vs $H_A:\mu>0.3$.

```{r}
# Data
y <- c(.593,.142,.329,.691,.231,.793,.519,.392,.418)

# Compute statistics
n <- length(y)
ybar <- mean(y); s <- sd(y)
tc <- (ybar-0.3)/(s/sqrt(n))

# Critical Region
critical_region <- qt(p=1-0.01, df=n-1)
tc > critical_region

# P-value
p_value <- pt(q=tc, df=n-1, lower.tail = F)
p_value
```


::: frame
## Type I and Type II Errors

All statistical tests have a level of uncertainty. We can quantify this through Type I and Type II errors. The probabilities of Type I and II error are $\alpha$ and $\beta$, respectively.
:::

::: frame
## Type I Error and Type II Error

- The experimenter can control only the Type I error probability. We do this by specifying an $\alpha$ for an experiment (before the data values are measured).

- When we are testing an hypothesis about $\mu$ using $\alpha$ for the test, the value of [Type II error probability $\beta$]{style="color: magenta"} depends on the actual value $\mu$ which is not known.

- This is because $\beta$ is the [probability of incorrectly failing to reject $H_0$ when $H_a$ is true]{style="color: magenta"}.

- When $H_a$ is true, the actual value of $\mu$ may be any value under $H_a$, i.e, a value not specified under $H_0$. Since this value of $\mu$ is unknown, $\beta$ cannot be calculated.
:::

## Illustration

::: frame
## Type I Error and Type II Error

- The implication of this is that when, based on a test, we find that we cannot reject $H_0$, *we will not say that we accept $H_0$*.

- Because if we say we accept $H_0$ and $\beta$ turns out to be large, then the probability is large that we will be committing a Type II error.

- Thus the correct way to state the decision is to say, that [we fail to reject $H_0$]{style="color: magenta"}.

:::



::: frame
## Power of the Test

- By definition, power of a test is the probability of rejecting $H_0$ for a specified value of $\mu$ under $H_a$.

- In practice, tests are designed to have large power for some $\mu$ values of interest so that they have small Type II error probabilities.

- We can relate to this idea by thinking of a test as having very good power if it has [a very good chance of detecting whether a change in $\mu$ has actually occurred]{style="color: magenta"}.

- This is usually done by selecting a sample size $n$ to be used for the experiment so that the desired power is achieved for a specified $\mu$ and $\alpha$.
:::





::: frame
## Departures from Normality

- When $\bar{Y}$ is a Normal random variable, i.e., when sampling from a Normal population with mean $\mu_0$, $\frac{\bar{Y}-\mu_0}{S/\sqrt{n}}$ is a $T_{n-1}$ random variable.

- This is a theoretical fact. In practice, however, we never sample exactly from a Normal population, so ($\bar{Y}- \mu_0)/S/\sqrt{n}$ will be only approximately $T_{n-1}$.

- How much effect can this have on C.I.'s and tests we construct?

- For symmetrically distributed populations and not too small $n$ there is little to worry about.

- For highly skewed population distributions the approximation can be terrible especially for small $n$.
:::

:::: frame
## 

- It is recommended that one look at a boxplot, Normal plot, and/or other graphics to see whether severe skewness of the sampling population is indicated. If not, proceed to use the $t$-distribution. If yes, one can use a nonparametric procedure or use a transformation. We will look at some of these later.
::::

::: frame
## Using Confidence Intervals to Test Hypotheses

- We can always look at a (1-$\alpha$)100% confidence interval and see what the result of a test would be if we carry out the test.

- For example, consider the (1-$\alpha$)100% confidence interval for $\mu$ $$\left(\bar{y} - t_{\alpha/2} \, \frac{s}{\sqrt{n}}, \quad
          \bar{y} + t_{\alpha/2} \, \frac{s}{\sqrt{n}}\right)$$

- Suppose that $\mu_0$ is [not included]{style="color: blue"} in the above confidence interval because $$\bar{y} - t_{\alpha/2} \, \frac{s}{\sqrt{n}}>\mu_0$$
:::

::: frame
## Using Confidence Intervals to Test Hypotheses - cont.

- By rearranging this we see that this is equivalent to $t_c$ being in the rejection region, i.e.,\
  $$\frac{\bar{y}-\mu_0}{s/\sqrt{n}} > t_{\alpha/2,(n-1)}$$

- Observe carefully that this is the same rejection region for the test of $H_0: \mu \leq  \mu_0$ vs. $H_a: \mu >
  \mu_0$ at level $\alpha/2$.

- That is, we will be rejecting $H_0$ [ at level $\alpha/2$]{style="color: magenta"} if\
  $$\frac{\bar{y}-\mu_0}{s/\sqrt{n}} > t_{\alpha/2,(n-1)}$$

- This is equivalent to saying that [if $\mu_0$ was not included in a 100(1-$\alpha$)% interval, then $H_0$ will be rejected if the test is carried out at $\alpha/2$ level]{style="color: magenta"}.
:::

::: frame
## Using Confidence Intervals to Test Hypotheses - cont.

- For [ a two-tailed test]{style="color: magenta"}, the confidence interval should be based on the same $\alpha$ as the test to make this inference.

[**In summary,**]{style="color: magenta"}

- To test $H_0: \mu \leq \mu_0, \quad H_a: \mu > \mu_0$, at level $\alpha/2$, use a 100(1-$\alpha$)% confidence interval.

- To test $H_0: \mu \geq \mu_0, \quad H_a: \mu < \mu_0$, at level $\alpha/2$, use a 100(1-$\alpha$)% confidence interval.

- $H_0: \mu = \mu_0, \quad H_a: \mu \neq \mu_0$, at level $\alpha$, use a 100(1-$\alpha$)% confidence interval.
:::
